Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-02 manuscript.tex
Finished analysis sections
New to revise the text  later.

\righthead{Tesseroids}
\ms{Submission} % manuscript number

@@

We have implemented the calculation of
the tesseroid gravitational fields
with adaptive discretization
in the open-source package \textit{Tesseroids}.
It is freely available
online through the software website\footnote{
\href{http://tesseroids.leouieda.com}{http://tesseroids.leouieda.com}}
under the BSD 3-clause open-source license.
An archived version of the source code
is available as part of this article
and online\footnote{\href{}{\todo[inline]{insert a ZENODO doi}}}.

@@

\subsection{Reproducing analysis and results}

All figures and numerical experiments in this article
were produced in IPython notebooks\footnote{
\href{http://ipython.org}{http://ipython.org}}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
We used \textit{pandas} by \citet{Mckinney2010}
to perform the data analysis,
\textit{matplotlib} by \citet{Hunter2007} for graphs,
and \textit{Mayavi} by \citet{Ramachandran2011} for 3D graphs.

The notebook files,
as well as instructions for installing and running the analysis,
are distributed along with the source code archive
that accompanies this article.
Alternatively,
all accompanying material can be downloaded
from an online source code repository\footnote{
\href{https://github.com/pinga-lab/paper-tesseroids}{https://github.com/pinga-lab/paper-tesseroids}}
or a persistent archive\footnote{
\href{}{\todo[inline]{Replace with the Zenodo/figshare DOI for the repo.}}}.

@@

The key controlling point of the adaptive discretization algorithm
is the distance-size ratio $D$ (equation~\ref{eq:condition}).

The specific value chosen for $D$ determines how many divisions will be made
(Figure~\ref{fig:division}).

Thus, $D$ indirectly controls both the accuracy of the integration
and the computation time.

In this section, we investigate the relationship between
values of the distance-size ratio and the integration error.

We perform the analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

The reference against which we compare the computed tesseroid fields
is a homogeneous spherical shell.

The shell has analytical solutions along the polar axis
\citep{Grombein2013}
and can be perfectly discretized into tesseroids.

We chose a spherical shell with a thickness of 1 km,
density of $2670\ kg.m^{-3}$,
bottom at height 0 km above the reference sphere,
and top at 1 km height.

We produced tesseroid models of the shell by discretizing it along the
horizontal dimensions into a regular mesh.

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.

Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.

Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points.

Fortunately, the symmetry of the shell allows us to consider the computation
point at any geographic coordinate.

Therefore, the effect of the shell will be same along the entire grid.

We compute the differences between the effects of the shell and the tesseroid
model on the grid and consider only the largest error.

By placing the grid on top of a particular tesseroid, we increase the chances
of capturing the true largest integration error.

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.

We repeated the analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.

Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

Figure~\ref{fig:dist-size-curves} shows
the maximum difference between the shell and tesseroid fields
as a function of $D$ for the four numerical experiments.

The differences are given as a percentage of the shell value.

We established a maximum tolerated error of $0.1\%$, represented by the
horizontal solid lines in Figure~\ref{fig:dist-size-curves}.

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.

For $g_z$, the same is achieved with $D=1.5$.

Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.

For a computation height of 260 km, the error curve for $G_{zz}$
intercepts the error threshold line at $D=2.5$.

This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.

To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.

Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.

Notice that the distance-size ratio needed to achieve $0.1\%$ accuracy
decreases as the computation height increases.

For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.

One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

We have implemented the values of the distance-size ratio producing $0.1\%$
accuracy as defaults for the software \textit{Tesseroids}.

We chose the conservative value of $D=8$ for the gravity gradient components
as a fail-safe alternative.

Users can control the value of $D$ used in the computations through
command-line arguments to achieve greater performance as a trade-off for
accuracy.

>>>>

\righthead{Tesseroids: modeling in spherical coordinates}
\ms{Submission} % manuscript number

@@

We have implemented the calculation of
the tesseroid gravitational fields
with adaptive discretization
in the open-source package \textit{Tesseroids}.
It is freely available
online through the software website\footnote{
\href{http://tesseroids.leouieda.com}{http://tesseroids.leouieda.com}}
under the BSD 3-clause open-source license.
An archived version of the source code
is available as part of this article
and online\footnote{
\href{}{\todo[inline]{Replace with the Zenodo DOI for the Tesseroids
source code.}}}.

@@

\subsection{Reproducing the analysis and results}

All figures and numerical experiments in this article
were produced in IPython notebooks\footnote{
\href{http://ipython.org}{http://ipython.org}}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
We used the libraries
\textit{pandas} by \citet{Mckinney2010}
to perform the data analysis,
\textit{matplotlib} by \citet{Hunter2007} for graphs,
and \textit{Mayavi} by \citet{Ramachandran2011} for 3D graphs.

The notebook files,
as well as instructions for installing and running the analysis,
are distributed along with the source code archive
that accompanies this article.
Alternatively,
all accompanying material can be downloaded
from an online source code repository\footnote{
\href{https://github.com/pinga-lab/paper-tesseroids}{https://github.com/pinga-lab/paper-tesseroids}}
or a persistent archive\footnote{
\href{}{\todo[inline]{Replace with the Zenodo/figshare DOI for the repository.}}}.

@@

The key controlling point of the adaptive discretization algorithm
is the distance-size ratio $D$ (equation~\ref{eq:condition}).
The specific value chosen for $D$ determines how many divisions will be made
(Figure~\ref{fig:division}).
Thus, $D$ indirectly controls both the accuracy of the integration
and the computation time.
In this section, we investigate the relationship between
the distance-size ratio and the integration error.
We perform the analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

The reference against which we compare the computed tesseroid fields
is a homogeneous spherical shell.
The shell has analytical solutions along the polar axis
\citep{Grombein2013}
and can be perfectly discretized into tesseroids.
We chose a spherical shell with a thickness of 1 km,
density of $2670\ kg.m^{-3}$,
bottom at height 0 km above the reference sphere,
and top at 1 km height.
We produced tesseroid models of the shell by discretizing it along the
horizontal dimensions into a regular mesh.

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points.
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid and consider only the largest error.
By placing the grid on top of a particular tesseroid, we increase the chances
of capturing the true largest integration error.

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

Figure~\ref{fig:dist-size-curves} shows
the maximum difference between the shell and tesseroid fields
as a function of $D$ for the four experiments.
The differences are given as a percentage of the shell value.
We established a maximum tolerated error of $0.1\%$, represented by the
horizontal solid lines in Figure~\ref{fig:dist-size-curves}.
Only results for the gravitational potential, $g_z$, and $g_{zz}$ are shown.
The results for the other diagonal components of the gravity gradient tensor
are similar to $g_{zz}$.
Figures for these components can be found in the supplementary material
(see section "Reproducing the analysis and results").

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $G_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio needed to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

We have implemented the values of the distance-size ratio producing $0.1\%$
accuracy determined above as defaults for the software \textit{Tesseroids}.
We chose the conservative value of $D=8$ for the gravity gradient components
as a fail-safe alternative.
Users can control the value of $D$ used in the computations through
command-line arguments to achieve greater performance at the cost of
accuracy.
